### 2022-05-27 LaTeX/report.tex
comments on report

\def\NN{$\mathcal{N}~$}
\def\GG{$\mathcal{G}~$}
\def\VV{$\mathcal{V}~$}
\def\EE{$\mathcal{E}~$}

@@

\begin{document}
\newgeometry{margin=2cm}
\begin{titlepage}
    \begin{figure}[!h]
        \centering
        \includegraphics[height = .1\textwidth]{/Users/vincent/Documents/MinesNancy/logoartem.png}
        \hspace{1cm}
        \includegraphics[height = .1\textwidth]{img/logoloria.jpg}
        \hspace{1cm}
        \includegraphics[height = .1\textwidth]{img/logoUL.png}
    \end{figure}
    \vspace{3cm}

@@

@@

@@

@@

\begin{definition}[SCC]
    Let \GG = (\VV, \EE) be an directed graph.
$\mathcal{C} \subseteq \mathcal{V}$ is a strongly connected component (SCC) of \GG if:
\begin{equation*}
    \forall x, y \in \mathcal{C}, (x \Rightarrow^* y) \wedge (y \Rightarrow^* x)
\end{equation*}
\textit{i.e.} there is a path between every $x$ and $y$ in $\mathcal{C}$.
\BlankLine
$\mathcal{C}$ is maximal, or $\mathcal{C}$ is a maximal SCC of \GG if there is no other SCC containing $\mathcal{C}$, \textit{i.e.} if:
\begin{equation*}
    \forall \mathcal{X}, (\mathcal{C} \subseteq \mathcal{X}) \wedge (\forall x, y \in \mathcal{X}, (x \Rightarrow^* y) \wedge (y \Rightarrow^* x)) \Longrightarrow \mathcal{C} = \mathcal{X}
\end{equation*}
\end{definition}

@@

@@

@@

@@

    \caption{Sequential set-based SCC algorithm}
\end{algorithm}
\ednote{sm}{%
  The algorithm should also explain how $\mathcal{S}$ is initialized.
}

@@

@@

\begin{proof}
    The disjointness of all on-stack partial SCCs is given by lemma \ref{lemma:disjointness}. Nodes from $\textsc{Visited} \setminus \textsc{Explored}$ are in \texttt{R} because they are being processed. So, $\textsc{Live} \subseteq \texttt{R}$.\\
    By L.6-7 of Algorithm \ref{alg:seqsetbased}, $\textsc{Visited} \subseteq \texttt{R}$.\\
    L.9-10 ensure that no explored node is pushed in \texttt{R}.\\
    L.24-25 keep the invariant by unstacking explored nodes from \texttt{R}, so $\texttt{R} \cap \textsc{Explored} = \varnothing$. Thus, $\texttt{R} = \textsc{Visited} \setminus \textsc{Explored} = \textsc{Live}$.
\end{proof}

@@

@@

@@

@@

@@

@@

@@

\isa{
    \bblue{definition} unite :: "\generic{v} $\Rightarrow$ \generic{v} $\Rightarrow$ \generic{v} env $\Rightarrow$ \generic{v} env" \where{}\\
    $~~~$"\blue{unite} \green{v w e} $\equiv$\\
    $~~~~~~$\bblue{let} \green{pfx} = takeWhile ($\lambda$\green{x}. \green{w} $\notin$ $\mathcal{S}$ \green{e x}) (stack \green{e});\\
    $~~~~~~~~~~$\green{sfx} = dropWhile ($\lambda$\green{x}. \green{w} $\notin$ $\mathcal{S}$ \green{e x}) (stack \green{e});\\
    $~~~~~~~~~~$\green{cc} = $\bigcup$ \{$\mathcal{S}$ \green{e x} $|$ \green{x}. \green{x} $\in$ set \green{pfx} $\cup$ \{hd \green{sfx}\}\}\\
    $~~~~~~$\bblue{in}  \green{e}\env{$\mathcal{S}$ := $\lambda$\green{x}. \bblue{if} \green{x} $\in$ \green{xx} \bblue{then} \green{cc} \bblue{else} $\mathcal{S}$ \green{e x}, stack := \green{sfx}}"
}

@@

The idea of this definition is to create a partition of \texttt{stack e = pfx @ sfx} such that \texttt{pfx} contains the nodes which are to be merged into \texttt{$\mathcal{S}$ e w} and \texttt{sfx} contains the root of \texttt{$\mathcal{S}$ e w} followed by the rest of the stack. Then, \texttt{cc} -- which stands for \textit{connected component} -- contains all the nodes which are equivalent to \texttt{w} in the sub-graph currently explored. The function \texttt{takeWhile} applied to a boolean function \texttt{P}\footnote{\texttt{P :: \generic{a} $\Rightarrow$ bool}} seen as a property and a list \texttt{xs} returns the elements of \texttt{xs} which satisfy \texttt{P} and stops to the first element not satisfying P. The function \texttt{dropWhile} is the opposite of \texttt{takeWhile}. Both the imperative and recursive versions of \texttt{unite} are equivalent.

\subsubsection{Ordering relation}\label{sec:orderingrelation}
In the proof, a precedence relation\footnote{In fact, a total order is being defined on stacks.} noted {\color{isa_dark_blue}$\bullet~\preceq~\bullet~\text{in}~\bullet$} will be needed on the stack. Let $x$ and $y$ be two nodes and $R$ be a stack. Informally, $x$ precedes $y$ in $R$ if $y$ was pushed in $R$ before $x$ (see \textsc{Figure} \ref{fig:stackorder}).

@@

All the different properties (\textit{i.e.} lemmas) which follow this definition in the Isabelle implementation are detailed in the natural mathematical writing in \nameref{appendix}. The right part of the notation represents the orders of priority for each operand since $\preceq$ is an infix operator.

@@

Now that the environment is set up, the actual algorithm -- seen as a function -- can be implemented.
Since Isabelle does not support loops, the implementation will be split into two mutually recursive functions. The main function is called \texttt{dfs} and takes its name after the Depth First Search algorithm because the algorithm \ref{alg:seqsetbased} roughly consists in a deep traversal of a graph. The second function is called \texttt{dfss} and represents the \textit{while} loop of the algorithm \ref{alg:seqsetbased}. The two functions are mutually recursive because they recursively call each other. In particular, \texttt{dfss} will call either itself or \texttt{dfs}, depending on the case.

@@

@@

\subsection{Formal proof}
\subsubsection{General scheme}
As the algorithm is composed of two mutually recursive functions, the correctness of the algorithm is proved by induction on the environment structure (cf \ref{envdef}). Since both \texttt{dfs} and \texttt{dfss} are quite complex, the proof is split into several parts. The idea is to prove for each function that its execution given some pre-conditions on the input environment implies some post-conditions on the output environment. Then, it has to be made for the mutually recursive calls as well, so that given the same pre-conditions on one function, the pre-conditions on the other function are also satisfied. Finally, it has to be proved that if the pre-conditions are satisfied for one function, and if the pre-conditions implie the post-conditions on the other function, then the post-conditions are also satisfied for the first function.

@@

Let us take a closer look to this invariant, taken in the same order as the definition above:
\begin{itemize}
    \item[$\bullet$] First, the stack is a set of distinct elements.
    \item[$\bullet$] All elements of the stack are visited.
    \item[$\bullet$] The set of explored nodes is a subset of the set of visited nodes.
    \item[$\bullet$] Explored nodes cannot be in the stack.
    \item[$\bullet$] The three next properties are about the equivalence relation $\mathcal{S}$.
    \item[$\bullet$] The union of the sets of equivalent nodes in the stack is equal to the set of visited nodes minus the set of explored nodes.
    \item[$\bullet$] A node in the stack can reach all nodes before it in the stack (\textit{i.e.} pushed after)
    \item[$\bullet$] $\mathcal{S}$ represents a set of strongly connected components (not maximal).
    \item[$\bullet$] For all explored nodes, the sub-graph induced by their successors is totally explored
    \item[$\bullet$] \texttt{sccs} is a set of maximal SCCs
\end{itemize}
\noindent
These properties are natural and most of them are easy to prove.

@@

It is also useful to induce a notion of monotonicity on the environments during the execution of the algorithm. This is define as follows through the definition of an ordering relation on environments:

@@

@@

\isa{
\bblue{definition} post\_dfs \where\\ 
"\blue{post\_dfs} \green{v prev\_e e} $\equiv$\\
$~~~~$wf\_env \green{e}\\
$~~\wedge$ ($\forall$ \green{x}. reachable \green{v} \green{x} $\longrightarrow$ \green{x} $\in$ visited e)\\
$~~\wedge$ sub\_env \green{prev\_e} \green{e}\\
$~~\wedge$ ($\forall$ \green{n} $\in$ set (stack \green{e}). reachable \green{n} \green{v})\\
$~~\wedge$ ($\exists$ \green{ns}. stack \green{prev\_e} = ns \@ (stack \green{e}))\\
$~~\wedge$ ($\forall$ \green{m n}. $\stackprec{m}{n}{prec\_e}$ $\longrightarrow$ \\ $~~~~~~$ ($\forall$ \green{u} $\in$ $\mathcal{S}$ \green{prev\_e m}. reachable \green{u} \green{v} $\wedge$ reachable \green{v} \green{n} $\longrightarrow$ $\mathcal{S}$ \green{e m} = $\mathcal{S}$ \green{e n}))\\
$~~\wedge$ ((\green{v} $\in$ explored \green{e} $\wedge$ stack \green{e} = stack \green{prev\_e}) $\lor$ \green{v} $\in$ $\mathcal{S}$ \green{e} (hd (stack e)))"
}

@@

>>>>

% sm: better: \def\NN{\ensuremath{\mathcal{N}}} etc.
\def\NN{$\mathcal{N}~$}
\def\GG{$\mathcal{G}~$}
\def\VV{$\mathcal{V}~$}
\def\EE{$\mathcal{E}~$}

@@

\begin{document}
\newgeometry{margin=2cm}
\begin{titlepage}
    \begin{figure}[!h]
        \centering
        \includegraphics[height = .1\textwidth]{img/logoartem.png}
        \hspace{1cm}
        \includegraphics[height = .1\textwidth]{img/logoloria.jpg}
        \hspace{1cm}
        \includegraphics[height = .1\textwidth]{img/logoUL.png}
    \end{figure}
    \vspace{3cm}

@@

\ednote{sm}{The introduction should also give an (informal) overview of what this report is about. Alternatively, split it into a section on the context of the work and one that introduces the subject. Avoid jumping directly into formal definitions without an introduction for the mere mortal.}

@@

\ednote{sm}{Again, the question whether or not to break pages at sections is decided by the \LaTeX{} style. I would not impose a page break here.}

@@

\ednote{sm}{The first definition should introduce directed graphs. In particular, it is not formally clear what ``\VV'', ``\GG'', and ``reachability'' refer to in the following definition.

Side note: Your macros \VV, \GG, \EE introduce trailing blank space, which they should not. (Also, they do not work within math environments, which is mildly annoying.)
You may want to use the xspace style for intelligently appending a space after a macro.}

@@

\begin{definition}[SCC]
    Let \GG = (\VV, \EE) be a directed graph.
$\mathcal{C} \subseteq \mathcal{V}$ is a strongly connected component (SCC) of \GG if:
\begin{equation*}
    \forall x, y \in \mathcal{C}, (x \Rightarrow^* y) \wedge (y \Rightarrow^* x)
\end{equation*}
\textit{i.e.} there is a path between every $x$ and $y$ in $\mathcal{C}$.
\BlankLine
$\mathcal{C}$ is maximal, or $\mathcal{C}$ is a maximal SCC of \GG if there is no other SCC containing $\mathcal{C}$, \textit{i.e.} if:
\begin{equation*}
    \forall \mathcal{X}, (\mathcal{C} \subseteq \mathcal{X}) \wedge (\forall x, y \in \mathcal{X}, (x \Rightarrow^* y) \wedge (y \Rightarrow^* x)) \Longrightarrow \mathcal{C} = \mathcal{X}
\end{equation*}
\end{definition}

@@

\ednote{sm}{This section should contain some text in order to explain the figures, beyond the captions. Also, Fig.~1 is more about connected graphs than SCCs.}

@@

\ednote{sm}{I don't quite understand this remark: $\mathcal{S}$ approximates mutual reachability, but doesn't capture it initially. Shouldn't the equivalence relation rather be $x \in \mathcal{S}(y) \land y \in \mathcal{S}(x)$?}

@@

\ednote{sm}{This definition is really about graphs and would fit better in section 2. Also, what you really want is $\textsc{Post}(v) = \{w : (v,w) \in \mathcal{E}\}$. The definition only provides an inclusion, for example $\textsc{Post}(v) = \emptyset$ satisfies it.}

@@

    \caption{Sequential set-based SCC algorithm}
\end{algorithm}
\ednote{sm}{%
  Give some more context than just the reference for the algorithm.
  The algorithm should also explain how $\mathcal{S}$ is initialized.
  And the result is really such that each element of \texttt{SCCs} is a (maximal) strongly connected component, not a set of components, and it is the set of SCCs in the subgraph of \GG reachable from $v_0$.
}

@@

\ednote{sm}{I do not find this proof convincing. The \texttt{while} loop is only executed at a certain point during the algorithm, and its condition is for two sets $\mathcal{S}$ to be different, not equal, as in the condition in the proof?}

@@

\begin{proof}
    The disjointness of all on-stack partial SCCs is given by lemma \ref{lemma:disjointness}. Nodes from $\textsc{Visited} \setminus \textsc{Explored}$ are in \texttt{R}\edmargin{sm}{not true: they have a (unique) representative in \texttt{R} but need not themselves be in \texttt{R}} because they are being processed. So, $\textsc{Live} \subseteq \texttt{R}$.\\
    By L.6-7 of Algorithm \ref{alg:seqsetbased}, $\textsc{Visited} \subseteq \texttt{R}$.\\
    L.9-10 ensure that no explored node is pushed in \texttt{R}.\\
    L.24-25 keep the invariant by unstacking explored nodes from \texttt{R}, so $\texttt{R} \cap \textsc{Explored} = \varnothing$. Thus, $\texttt{R} = \textsc{Visited} \setminus \textsc{Explored} = \textsc{Live}$.
\end{proof}

@@

\ednote{sm}{Should be strengthened to ``the set of maximal SCCs reachable from $v_0$.}

\ednote{sm}{A figure illustrating the execution of the algorithm would be nice, also in the talk presenting the project.}

@@

\ednote{sm}{I don't really understand the prose above. A recursive function definition has to be based on some underlying inductive definition, which is simply not available for graphs. You say this, but the beginning of the paragraph seems to say otherwise. I'd just say that Isabelle provides a construction for inductive predicate definitions, which is appropriate here, and then explain that the two clauses represent the reflexive case and the extension of reachability by prepending an edge.}

@@

\ednote{sm}{I'm afraid that the blue color used for \texttt{reachable} here can be confusing, although Isabelle indeed displays the definition in this way. I'd make it black throughout.}

@@

\ednote{sm}{Note that this is the ``mirror'' of clause \texttt{reachable\_succ} (appending an edge).}

@@

\ednote{sm}{``Telling Isabelle'' sounds like you are introducing an axiom. Rather, these facts are deduced using Isabelle from the above definitions.}

@@

\ednote{sm}{Note that this is different from the \textsc{Unite} function introduced earlier, and perhaps rename one of the two. Also, I'd introduce this function together with the \texttt{dfs} and \texttt{dfss} functions where it is used. Finally, the heading of the subsection doesn't seem to fit the text.}

@@

\isa{
    \bblue{definition} unite :: "\generic{v} $\Rightarrow$ \generic{v} $\Rightarrow$ \generic{v} env $\Rightarrow$ \generic{v} env" \where{}\\
    $~~~$"\blue{unite} \green{v w e} $\equiv$\\
    $~~~~~~$\bblue{let} \green{pfx} = takeWhile ($\lambda$\green{x}. \green{w} $\notin$ $\mathcal{S}$ \green{e x}) (stack \green{e});\\
    $~~~~~~~~~~$\green{sfx} = dropWhile ($\lambda$\green{x}. \green{w} $\notin$ $\mathcal{S}$ \green{e x}) (stack \green{e});\\
    $~~~~~~~~~~$\green{cc} = $\bigcup$ \{$\mathcal{S}$ \green{e x} $|$ \green{x}. \green{x} $\in$ set \green{pfx} $\cup$ \{hd \green{sfx}\}\}\\
    $~~~~~~$\bblue{in}  \green{e}\env{$\mathcal{S}$ := $\lambda$\green{x}. \bblue{if} \green{x} $\in$ \green{cc} \bblue{then} \green{cc} \bblue{else} $\mathcal{S}$ \green{e x}, stack := \green{sfx}}"
}

@@

The idea of this definition is to create a partition of \texttt{stack e = pfx @ sfx} such that \texttt{pfx} contains the nodes which are to be merged into \texttt{$\mathcal{S}$ e w} and \texttt{sfx} contains the root of \texttt{$\mathcal{S}$ e w} followed by the rest of the stack. Then, \texttt{cc} -- which stands for \textit{connected component} -- contains all the nodes which are equivalent to \texttt{w} in the sub-graph currently explored. The function \texttt{takeWhile} applied to a boolean function \texttt{P}\footnote{\texttt{P :: \generic{a} $\Rightarrow$ bool}} seen as a property and a list \texttt{xs} returns the elements of \texttt{xs} which satisfy \texttt{P} and stops at the first element not satisfying \texttt{P}. The function \texttt{dropWhile} is the opposite of \texttt{takeWhile}. Both the imperative and recursive versions of \texttt{unite} are equivalent.

\ednote{sm}{The two are equivalent within the context of the algorithm, not in general. And the equivalence has not been proved. In fact, the non-recursive (why ``imperative''?) definition is intended to be simpler for the proof because it avoids introducing separate pre- and post-conditions for the function and proving such a ``contract''.}

\subsubsection{Ordering relation}\label{sec:orderingrelation}
In the proof, a precedence relation\footnote{In fact, a total order is being defined on stacks.}\edmargin{sm}{partial? don't have $x \preceq y \lor y \preceq x$} noted {\color{isa_dark_blue}$\bullet~\preceq~\bullet~\text{in}~\bullet$} will be needed on the stack. Let $x$ and $y$ be two nodes and $R$ be a stack. Informally, $x$ precedes $y$ in $R$ if $y$ was pushed in $R$ before $x$ (see \textsc{Figure} \ref{fig:stackorder}).

@@

All the different properties (\textit{i.e.} lemmas) which follow this definition in the Isabelle implementation are detailed in the natural mathematical writing in the \nameref{appendix}. The right part of the notation represents the orders of priority for each operand since $\preceq$ is an infix operator.

@@

Now that the environment is set up, the actual algorithm -- seen as a function -- can be implemented.
Since Isabelle does not support loops, the implementation will be split into two mutually recursive functions. The main function is called \texttt{dfs} and takes its name after the Depth First Search algorithm because the algorithm \ref{alg:seqsetbased} roughly consists in a deep traversal of a graph. The second function is called \texttt{dfss} and represents the \textit{while} loop\edmargin{sm}{\textbf{foreach} loop?!} of the algorithm \ref{alg:seqsetbased}. The two functions are mutually recursive because they recursively call each other. In particular, \texttt{dfss} will call either\edmargin{sm}{both itself and \texttt{dfs}} itself or \texttt{dfs}, depending on the case.

@@

\ednote{sm}{\texttt{force} finishes the proof of pattern completeness, the proof of termination remains open, and it would actually show that these are well-defined functions. Also, you should say that \texttt{force} is more aggressive \emph{than \texttt{auto}}.}

@@

\subsection{Formal proof}
\subsubsection{General scheme}
As the algorithm is composed of two mutually recursive functions, the correctness of the algorithm is proved by induction on the environment structure (cf \ref{envdef}).\edmargin{sm}{mutual induction on the functions? There is no induction principle for environments.} Since both \texttt{dfs} and \texttt{dfss} are quite complex, the proof is split into several parts. The idea is to prove for each function that its execution given some pre-conditions on the input environment implies some post-conditions on the output environment. Then, it has to be made for the mutually recursive calls as well, so that given the same pre-conditions on one function, the pre-conditions on the other function are also satisfied. Finally, it has to be proved that if the pre-conditions are satisfied for one function, and if the pre-conditions imply the post-conditions on the other function, then the post-conditions are also satisfied for the first function.

@@

Let us take a closer look to this invariant, taken in the same order as the definition above:
\begin{itemize}
    \item[$\bullet$] First, the stack is a list of distinct elements.
    \item[$\bullet$] All elements of the stack are visited.
    \item[$\bullet$] The set of explored nodes is a subset of the set of visited nodes.
    \item[$\bullet$] Explored nodes cannot be in the stack.
    \item[$\bullet$] The three next properties are about the equivalence relation $\mathcal{S}$.
    \item[$\bullet$] The union of the sets of equivalent nodes in the stack is equal to the set of visited nodes minus the set of explored nodes.
    \item[$\bullet$] A node in the stack can reach all nodes before it in the stack (\textit{i.e.} pushed later).
    \item[$\bullet$] $\mathcal{S}$ represents a set of strongly connected components (not maximal).
    \item[$\bullet$] For all explored nodes, the sub-graph induced by their successors is totally explored.
    \item[$\bullet$] \texttt{sccs} is a set of maximal SCCs
\end{itemize}
\noindent
These properties are natural and most of them are easy to prove.

\ednote{sm}{Actually, there is a bit of redundancy here. For example, the second and fourth conjunct follow from the fifth and eighth. This is not a bad thing per se since it may help automatic proof, but could be discussed.}

@@

It is also useful to induce a notion of monotonicity on the environments during the execution of the algorithm. This is defined as follows through the definition of an ordering relation on environments:

@@

\ednote{sm}{Add a bit of explanation, as for the basic invariant above, in particular for the last conjunct.}

@@

\isa{
\bblue{definition} post\_dfs \where\\ 
"\blue{post\_dfs} \green{v prev\_e e} $\equiv$\\
$~~~~$wf\_env \green{e}\\
$~~\wedge$ ($\forall$ \green{x}. reachable \green{v} \green{x} $\longrightarrow$ \green{x} $\in$ visited e)\\
$~~\wedge$ sub\_env \green{prev\_e} \green{e}\\
$~~\wedge$ ($\forall$ \green{n} $\in$ set (stack \green{e}). reachable \green{n} \green{v})\\
$~~\wedge$ ($\exists$ \green{ns}. stack \green{prev\_e} = ns @ (stack \green{e}))\\
$~~\wedge$ ($\forall$ \green{m n}. $\stackprec{m}{n}{prec\_e}$ $\longrightarrow$ \\ $~~~~~~$ ($\forall$ \green{u} $\in$ $\mathcal{S}$ \green{prev\_e m}. reachable \green{u} \green{v} $\wedge$ reachable \green{v} \green{n} $\longrightarrow$ $\mathcal{S}$ \green{e m} = $\mathcal{S}$ \green{e n}))\\
$~~\wedge$ ((\green{v} $\in$ explored \green{e} $\wedge$ stack \green{e} = stack \green{prev\_e}) $\lor$ \green{v} $\in$ $\mathcal{S}$ \green{e} (hd (stack e)))"
}

\ednote{sm}{Add some explanations, and continue \dots}

\ednote{sm}{Don't forget to write a conclusion, explaining what has been done, what is missing / could be improved, and what your experience has been.}

@@

\ednote{sm}{Add a few more references, such as about Isabelle, and possibly other SCC algorithms (Dijkstra, Tarjan). For each reference, indicate where it was published, not just the year.}
